Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-14 ieee_paper.tex
Added that the size was in bytes in the tables. ` I checked Eric's formulas. They are correct. Changed some english.

\begin{figure}
  \begin{center}
    \scalebox{.9}{
      \begin{tabular}{|c|l|}
        %\hline
        %& & \\
        \hline
        $b_i$ & number of bytes per index \\
        $b_x$ & number of bytes per value \\
        $n_z$ & number of nonzeros per row $A$ \\
        $n_r$ & number of column/rows of $A$ \\
        $n_c$ & total number of non-zero\\
        $n_v$ & number of {\tt x} vectors \\
        $n_m$ & number of matrices \\
        $s_M$ & size of the matrice(s)\\
        $s_x$ & size of the {\tt x} vector(s)\\
        $s_y$ & size of the {\tt y} vector(s)\\
        \hline
        $cl$    & size of a cache line \\
        $b_{wT}$ & number of bytes written to memory  \\
        $b_{rT}$ & minimum number of bytes read from memory  \\
        $b_T$   & minimum number of bytes transferred  \\
        $B_{rT}$ & maximum number of bytes read from memory \\
        $B_T$   & maximum number of bytes transferred  \\
        $O$     & number of floating point operations \\
        $I_b$   & maximal computational intensity\\
        $I_w$   & minimum computational intensity\\        
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{Notations}
  \label{tab:not}
\end{figure}

@@

The matrix is composed of $n_r$ rows and columns with $n_z$ non-zero
per row leading to a total of $$n_c = n_r n_z$$ non-zero entries in
the matrix. Each of these non-zero entry have one index of size $b_i$
and $n_m$ values of size $b_x$. The matrices have a total size
of $$s_M = n_c (b_i + b_x n_m) = n_r n_z (b_i + b_x n_m)$$

If we assume an algorithm where the rows are processed the one after
the one, the amount of memory that need to be written is exactly the
size of the {\tt y} vector. (This assumption removes the possibility of
cache partitioning techniques.) $$b_{wT} = s_y = n_v n_m b_x n_r$$

The amount of data read from memory depends highly on how the
algorithm execute, how the matrix is structured. But in the best case
the whole matrix is read a single time and the whole {\tt x} vector is
read a single time from the main memory. (Assuming the whole vector
{\tt x} is used is reasonnable since only oddly shaped matrix will
left a dimension of the vector unused.) $$b_{rT} = s_M + s_x = n_r n_z
(b_i + b_x n_m) + n_v b_x n_r$$
 $$b_T = b_{rT} + b_{wT} =  n_r n_z (b_i + b_x n_m) + n_v b_x n_r (1 + n_m)$$

Notice that there is no reason for a piece of the matrix to be read
multiple times. But assuming that each element of the {\tt x} vector
is read a single time is a strong assumption. If using a single core,
it assumes that either the cache of the architecture can store all the
{\tt x} vector or that the matrix is sufficiently well structured for
causing no cache trashing. If using multiple core, this assumes that
no element of the {\tt x} vectors will be used by multiple
cores. \cite{Saule12} showed that there is very little cache trashing
in practice, but it showed that having elements of the vectors used by
multiple cores can have a significant impact on the performance
(growing with $n_v$).

On the other hand, in the worst case, every time the {\tt x} vector is
accessed, the value needs to be transfered from memory again. So in
total, there are as many transfers as the number of non zero in the
matrix. Notice that most architectures do not allow to read the memory
one byte at a time but always in a set of multiple bytes. On most
cached architecture, that minimum size corresponds to the size $cl$ of
a cacheline (typically 64 bytes). Notice that when there are multiple
vectors, each non-zero element uses $n_v$ consecutive entries. The
worst case number of bytes read and transfered are 
$$B_{rT} = s_M + n_c cl \ceil{\frac{n_vb_x}{cl}} $$ 
$$B_T = n_v n_m b_x n_r + n_r n_z \left ( b_i + b_x n_m +  cl \ceil{\frac{n_vb_x}{cl}} \right)$$

@@

In the special case $n_v=n_m=1$ in single precision
$$I_b = \frac{1}{4} \frac{ n_z }{ n_z + 1} \approx \frac{1}{4}$$
$$I_w = \frac{n_z }{2  + 36  n_z} \approx \frac{1}{36}$$

In the special case $n_v=n_m=4$ in single precision
$$I_b = \frac{32 n_z }{ n_z 20 + 80} \approx \frac{32}{20} = \frac{8}{5} $$
$$I_w = \frac{32 n_z }{64 + n_z 84} \approx \frac{32}{84} = \frac{8}{21} $$

>>>>

\begin{figure}
  \begin{center}
    \scalebox{.9}{
      \begin{tabular}{|c|l|}
        %\hline
        %& & \\
        \hline
        $b_i$ & number of bytes per index \\
        $b_x$ & number of bytes per value \\
        $n_z$ & number of nonzeros per row $A$ \\
        $n_r$ & number of column/rows of $A$ \\
        $n_c$ & total number of non-zero\\
        $n_v$ & number of {\tt x} vectors \\
        $n_m$ & number of matrices \\
        $s_M$ & size of the matrice(s) in bytes\\
        $s_x$ & size of the {\tt x} vector(s) in bytes\\
        $s_y$ & size of the {\tt y} vector(s) in bytes\\
        \hline
        $cl$    & size of a cache line in bytes\\
        $b_{wT}$ & number of bytes written to memory  \\
        $b_{rT}$ & minimum number of bytes read from memory  \\
        $b_T$   & minimum number of bytes transferred  \\
        $B_{rT}$ & maximum number of bytes read from memory \\
        $B_T$   & maximum number of bytes transferred  \\
        $O$     & number of floating point operations \\
        $I_b$   & maximal computational intensity\\
        $I_w$   & minimum computational intensity\\        
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{Notations}
  \label{tab:not}
\end{figure}

@@

The matrix is composed of $n_r$ rows and columns with $n_z$ non-zeros
per row leading to a total of $$n_c = n_r n_z$$ non-zero entries in
the matrix. Each of these non-zero entries has one index of size $b_i$
and $n_m$ values of size $b_x$. The matrices have a total size
of $$s_M = n_c (b_i + b_x n_m) = n_r n_z (b_i + b_x n_m)$$

If we assume an algorithm where the rows are processed one after
the other, the amount of memory written is precisely the
size of the {\tt y} vector. (This assumption removes the possibility of
cache partitioning techniques.) $$b_{wT} = s_y = n_v n_m b_x n_r$$

The amount of data read from memory depends highly on both the algorithm's
execution path, and on %algorithm execute, 
how the matrix is structured. But in the best case
both the matrix $A$ %is read a single time 
and the source vector {\tt x} are read once
%read a single time 
from the main memory. (We assume that all the elements of $\tt x$ are involved
in the SpMV.)  %is used is reasonnable since only oddly shaped matrix will
%left a dimension of the vector unused.) 
Thus, $$b_{rT} = s_M + s_x = n_r n_z
(b_i + b_x n_m) + n_v b_x n_r$$
 $$b_T = b_{rT} + b_{wT} =  n_r n_z (b_i + b_x n_m) + n_v b_x n_r (1 + n_m)$$

Notice that there is no reason for a piece of the matrix to be read
multiple times. But assuming that each element of the {\tt x} vector
is read a single time is a strong assumption. If using a single core,
it assumes that either the cache of the architecture can store the full
{\tt x} vector or that the matrix is sufficiently well structured 
causing no cache trashing. If using multiple core, this assumes that
no element of the {\tt x} vectors will be used by multiple
cores. \cite{Saule12} showed that there is very little cache trashing
in practice, but it showed that having elements of the vectors used by
multiple cores can have a significant impact on the performance
(growing with $n_v$).

On the other hand, in the worst case, every time the {\tt x} vector is
accessed, the value needs to be transfered from memory again. So in
total, there are as many transfers as the number of non-zeros in the
matrix. Note however that most architectures cannot read memory a single 
byte at a time. Instead, a minimum number of bytes,equal to the size o a cacheline, 
are transfered at once. 
%multiple bytes, with each chunk equal to the data do not allow to read the memory
%one byte at a time but always in a set of multiple bytes. On most
%cached architecture, that 
%the minimum size corresponds to the size $cl$ of
%a cacheline (typically 64 bytes). 
When there are multiple
vectors, each non-zero element uses $n_v$ consecutive entries. The
worst case number of bytes read and transfered are 
$$B_{rT} = s_M + n_c cl \ceil{\frac{n_vb_x}{cl}} $$ 
$$B_T = n_v n_m b_x n_r + n_r n_z \left ( b_i + b_x n_m +  cl \ceil{\frac{n_vb_x}{cl}} \right)$$

@@

In the special case $n_v=n_m=1$ in single precision
$$I_b = \frac{1}{4} \frac{ n_z }{ n_z + 1} \approx \frac{1}{4}$$
%$$I_w = \frac{n_z }{2  + 36  n_z} \approx \frac{1}{36}$$
$$I_w = \frac{n_z }{2  + (4+cl/2)  n_z} \approx \frac{1}{4+cl/2}$$

In the special case $n_v=n_m=4$ in single precision
$$I_b = \frac{32 n_z }{ n_z 20 + 80} \approx \frac{32}{20} = \frac{8}{5} $$
%$$I_w = \frac{32 n_z }{64 + n_z 84} \approx \frac{32}{84} = \frac{8}{21} $$
$$I_w = \frac{32 n_z }{64 + (20+cl) n_z} \approx \frac{32}{84} = \frac{8}{21} $$
The above formulas have been applied to MIC, which has a cache size $cl=64$ bytes.
